Formalization and Automation of Variants of G\"{o}del's Ontological Proof of God's Existence

### 2013-08-16 Abstract/GodProof-Abstract.tex
added Bibliography.bib and slightly modified abstract

\title{
  Formalizations of G\"{o}del's Ontological Proof of \\
  God's Existence
  %\thanks{Supported by ... }
}

@@

\institute{
  Theory and Logic Group, Vienna University of Technology, Austria \\
  \email{bruno@logic.at}
  \and
  Dahlem Center for Intelligent Systems, Freie Universit\"{a}t Berlin, Germany\\
  \email{c.benzmueller@gmail.com}
}

@@

Attempts to prove the existence (or non-existence) of God by means of 
abstract ontological arguments are an old tradition in philosophy and theology,
G\"{o}del's proof is a modern culmination of this tradition, following particularly 
the footsteps of Leibniz.
%
G\"{o}del defines God as a being who possesses all \emph{positive} properties.
He does not extensively discuss what positive properties are, 
but instead he states a few reasonable but debatable axioms that they should satisfy.
From these axioms, he claims (without detailed proof) that:
\begin{enumerate}
\item If it is possible that God exists, then it is necessary that God exists.
\item It is possible that God exists.
\end{enumerate}

In the literature \cite{ToDo} slightly different versions of the axioms exist 
and derivations of the above mentioned claims are presented with various degrees
of detail and formality. Here \cite{ToDo: GitHub Repository} (a version of) G\"{o}del's proof is 
constructed for the first-time
with the utmost degree of detail and formality. The following has been done:
\begin{itemize}
\item A detailed natural deduction proof.
%
\item A formalization of the axioms and theorems in the TPTP THF format \cite{ToDo}.
%
\item Automatic verification of the consistency of the axioms and 
definitions with ToDo \cite{ToDo}.
%
\item Automatic demonstration of the theorems with the provers LEO-II \cite{ToDo} and Satallax \cite{ToDo}.

@@

\item A formalization using the Isabelle proof assistant \cite{ToDo} partially automated with Sledgehammer \cite{ToDo} and Metis \cite{ToDo}.
\end{itemize}

G\"{o}del's proof is challenging to formalize and automatically verify 
because it requires very expressive logical languages with 
modal operators (\emph{possible} and \emph{necessary}) and higher-order quantifiers. 
The computer-assisted formalizations rely on an embedding of the modal logic S5 
into classical higher-order logic with Henkin 
semantics \cite{ToDo: Current works of Christoph Benzmüller and Larry Paulson} and 
employed recently developed interactive and automated deduction tools designed for this logic.

This work attests the maturity of contemporary interactive and 
automated deduction tools and 
opens new perspectives for a computer-assisted theoretical philosophy. 
The critical discussion of the underlying 
concepts, definitions and axioms remains a human responsibility, but the computer can assist 
in building and checking rigorously correct logical arguments. In case of 
logico-philosophical disputes, the computer can check the disputing arguments
and partially fulfill Leibniz' dictum:
Calculemus --- Let us calculate!
\end{document}

>>>>

\title{
  Formalization and Automation of Variants of G\"{o}del's Ontological Proof of 
  God's Existence
  %\thanks{Supported by ... }
}

@@

\institute{
  Dahlem Center for Intelligent Systems, Freie Universit\"{a}t Berlin, Germany\\
  \email{c.benzmueller@gmail.com}
  \and 
  Theory and Logic Group, Vienna University of Technology, Austria \\
  \email{bruno@logic.at}
}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  G\"{o}del's proof \cite{Goedel} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
%
G\"{o}del defines God as a being who possesses all \emph{positive} properties.
He does not extensively discuss what positive properties are, 
but instead he states a few reasonable but debatable axioms that they should satisfy.
%
In our recent work we did employ the following axioms\footnote{Say where they come from and how they relate to G\"odel.}: (1) Any property strictly implied by a positive property is positive,
(2) A property is positive if and only if its negation is not positive, and (3) the property of being God-like is positive. Moreover, the following definitions are employed:
(A) x is God-like if and only if x incorporates all positive properties, (B) a property P is essential for x (and essence of x) if and only if P is a property of x and every property Q that x has is strictly implied by P, (C) x necessarily exists if and only if every essence of x is necessarily exemplified.

From these axioms and definitions we then infer:
\begin{itemize}
\item Positive properties are eventually exemplified.
\item Eventually God exists.
\item If x is God-like, then the property of being 
God-like is an essence of x.
\item Necessarily God exists.
\end{itemize}

% From these axioms, he claims (without detailed proof) that:
% \begin{enumerate}
% \item If it is possible that God exists, then it is necessary that God exists.
% \item It is possible that God exists.
% \end{enumerate}

In the literature \cite{ToDo} slightly different versions of the axioms and definitions exist 
and derivations of the above mentioned claims are presented with various degrees
of detail and formality. The above variant of G\"{o}del's proof has now been 
constructed for the first-time
with the utmost degree of detail and formality; cf.~\cite{ToDo:GitHubRepository}. The following has been done:
\begin{itemize}
\item A detailed natural deduction proof.
%
\item A formalization of the axioms and theorems in the TPTP THF format \cite{J22}.
%
\item Automatic verification of the consistency of the axioms and 
definitions with Nitpick \cite{Nitpick}.
%
\item Automatic demonstration of the theorems with the provers LEO-II \cite{LEO-II} and Satallax \cite{Satallax}.

@@

\item A formalization using the Isabelle proof assistant \cite{Isabelle} partially automated with Sledgehammer \cite{ToDo} and Metis \cite{ToDo}.
\end{itemize}

G\"{o}del's proof is challenging to formalize and verify 
because it requires very expressive logical languages with 
modal operators (\emph{possible} and \emph{necessary}) and higher-order quantifiers. 
Our computer-assisted formalizations rely on an embedding of the modal logic S5 
into classical higher-order logic with Henkin 
semantics \cite{J23} and 
employed recently developed interactive and automated deduction tools designed for this logic.

This work attests the maturity of contemporary interactive and
automated deduction tools and opens new perspectives for a
computer-assisted theoretical philosophy.  
The critical discussion of the underlying concepts, definitions and
axioms remains a human responsibility, but the computer can assist in
building and checking rigorously correct logical arguments. In case of
logico-philosophical disputes, the computer can check the disputing
arguments and partially fulfill Leibniz' dictum: Calculemus --- Let us
calculate!

Future work includes an
extensive study of other formalizations of ontological arguments with
our machinery.

\bibliographystyle{plain}
\bibliography{Bibliography}

\end{document}
